The Precambrian paleogeography of Laurentia

### 2020-11-18 Manuscript/Laurentia_Paleogeo_Manuscript.tex
implementing some edits from JS suggestions

The Grenvillian orogeny was a protracted interval of continent-continent collision (ca. 1.09 to 0.98 Ga) leading to amphibolite to granulite facies metamorphism throughout the orogen \citep{McLelland2010a}. Evidence of large-scale continent-continent collision at the time of the Ottawan Phase of the Grenvillian orogeny is recorded in Texas \citep{Grimes2004a}, through the Blue Ridge Appalachian inliers \citep{Johnson2020a}, through Ontario and to the Labrador Sea \citep{Rivers2008a}. The orogeny is interpreted to have resulted in the development of a thick plateau associated with the Ottawan orogenic phase (ca. 1090 to 1030 Ma; \citealp{Rivers2008a}). Continued convergence during the Rigolet phase of the Grenvillian orogeny led to the development of the Grenville Front tectonic zone and ended ca. 980 Ma \citep{Hynes2010a}.

@@

There is significantly less preserved Mesoproterozoic crustal growth on the western margin of Laurentia (Fig. \ref{fig:Laurentia_map}) and the tectonic history through the Mesoproterozoic Era is not as well elucidated as on the southern to eastern margin. There are thick successions of Paleoproterozoic sediments such as the ca. 1.66 to 1.62 Ga Wernecke Supergroup in Yukon, Canada that comprises $\sim$14 km of  fine-grained siliciclastic and carbonate rocks \citep{Furlanetto2016a}. This succession is interpreted to have resulted from rifting followed by passive margin thermal subsidence \citep{Furlanetto2016a}. These metasedimentary rocks were deformed and metamorphosed during the ca. 1.60 Ga Racklan orogeny associated with collision of an arc terrane and potentially a conjugate continent with associated orogenesis (referred to as the Forward orogeny) further into the craton \citep{Thorkelson2005a, Furlanetto2013a, Furlanetto2016a}. Further south along Laurentia's western margin, the 15 to 20 km thick package of sedimentary rocks of the Belt-Purcell Supergroup is associated with a ca. 1.47 to 1.40 rift \citep{Evans2000c}. While the rift is typically interpreted as being intracontinental \citep{Lydon2004a}, the tectonic setting in which it formed is debated. \citet{Hoffman1989c} proposed that it may be a remnant back-arc basin trapped within a continent, while others envision it as being associated with continental rifting along the margin associated with separation of a conjugate continent \citep{Jones2015a}. This region is interpreted to have been subsequently deformed during a ca. 1.37 to 1.33 Ga event known as the East Kootenay orogeny that is constrained by granite crystallization and authigenic monazite dates \citep{McMechan1982a, Nesheim2012a, McFarlane2015a}.

Taken together, this late Paleoproterozoic and Mesoproterozoic tectonic history provides significant constraints on paleogeographic reconstructions. In particular, the long-lived history of accretionary orogenesis along the southeast (present-day coordinates) of Laurentia from the initiation of the Yavapai orogeny (ca. 1.71 Ga) to the end of the Shawinigan orogeny (ca. 1.16 Ga) requires a long-lived open margin without a major conjugate continent until the time of terminal Grenvillian collisional orogenesis \citep{Karlstrom2001a}. This constraint is incorporated into paleogeographic models such as that of \citet{Zhang2012a} and \citet{Pehrsson2015a} which maintain a long-lived convergent margin throughout the Mesoproterozoic, but in some reconstructions other continental blocks are reconstructed into positions that are seemingly incompatible with this record of accretionary orogenesis (e.g. Amazonia in \citealp{Elming2009a, Elming2021a}). The high-grade metamorphism associated with the Ottawan phase of the Grenvillian orogeny itself strongly suggests a collision between Laurentia and another continent ca. 1080 Ma --- the geological observation of which first led to the formulation of the hypothesis of the supercontinent Rodinia \citep{Hoffman1991a}. This extensive and major collisional orogenic history recorded in Laurentia, and also present on other Proterozoic continents, at this time remains a strong piece of evidence that a supercontinent or (proto)supercontinent formed at the 1.0 Ga Mesoproterozoic to Neoproterozoic transition. Note that while the term Grenville orogeny or Grenville belt is used rather loosely throughout much of the literature to refer to any late Mesoproterozoic orogenic belt, the timeline of orogenesis on the Laurentia margin has more nuanced constraints than this usage. These constraints can be comparatively assessed when evaluating potential conjugate continents to Laurentia associated with the orogen (Fig. \ref{fig:tectonic_history}).

@@

The subsequent Neoproterozoic tectonic history of Laurentia is dominantly a record of rifting (Fig. \ref{fig:tectonic_history}). Along the western margin of Laurentia, by ca. 760 Ma there was rifting leading to deposition in basins from the Death Valley region of southwest Laurentia to the Mackenzie Mountains of northwest Laurentia \citep{Macdonald2013a, Dehler2017a, Rooney2017a}. The emplacement of the ca. 780 Ma Gunbarrel large igneous province \citep{Harlan2003a} along this margin and the subsequent extension recorded in the western Laurentia basins is commonly interpreted to be associated with the break-up of Laurentia and a conjugate continent to the western margin (often interpreted to be Australia; \citealp{Li2008a}). Extensional basin development continued into the Cryogenian Period with active normal faulting occurring during the deposition of both Sturtian (ca. 717 to 656 Ma) and Marinoan (ca. 645 to 635 Ma) glacial deposits in the Death Valley region of southwest Laurentia \citep{Nelson2020a}. Additionally, Cryogenian volcanics along the western Laurentia margin (e.g. \citealp{Eyster2018a}) are interpreted to be the result of active rifting.  A puzzlingly feature of this record of active rifting is that it significantly predates interpreted passive margin thermal subsidence closer to the ca. 539 Ma Neoproterozoic-Phanerozoic boundary that has been linked to lithospheric thinning \citep{Bond1984a, Levy1991a}. If the interpretation of a conjugate continent rifting of the margin prior to the ca. 717 Tonian-Cryogenian boundary is correct, it is unclear why there would be minimal thermal subsidence until the Ediacaran (after 635 Ma in \citealp{Levy1991a} and \citealp{Witkosky2018a}). While the geological evidence supports prolonged extensional tectonism along the western margin of Laurentia (a portion of which could be strike-slip and transtensional; \citealp{Smith2015b}), it suggests that there was significant lithospheric thinning associated with rifting later than the timing of rifting typically implemented in models of Rodinia break-up. This record of Neoproterozoic basin development led \cite{Yonkee2014a} to propose that the early ca. 780 Ma rifting was intracratonic and that while it may have led to some associated thermal subsidence that there was a second interval of rifting and thermal subsidence associated with Australia rifting away in the Ediacaran (later than in most models). Another possibility, along the lines of that proposed in \citet{Ross1991a} and \citet{Colpron2002a}, is that ca. 760 Ma extensional tectonism is an inboard record of rifting and passive margin development that occurred further to the west. In this model, subsequent continental rifting that drove lithospheric thinning, perhaps associated with the departure of a ribbon continent rather than an already departed major conjugate continent, would be the cause of Ediacaran to Cambrian thermal subsidence.

In northwest Laurentia from the Ogilvie Mountains of Yukon to Victoria Island, the sedimentary rock record is distinct from further south as it also records earlier Neoproterozoic basin development during the Tonian Period in addition to Cryogenian basin development \citep{Macdonald2012a}. Lithospheric extension is interpreted from basin development that accommodated deposition of the lower Fifteenmile Group with maximum depositional ages of ca. 1050 Ma with ongoing basin development ca. 812 Ma (age constraint from a U-Pb zircon date on a tuff within the upper Fifteenmile Group; \citealp{Macdonald2010a}) potentially through thermal subsidence \citep{Macdonald2012a}. Earlier basin development in the region recorded by the Mesoproterozoic/Neoproterozoic Pinguicula Group could provide valuable insight on tectonic history as it has been interpreted to have been deposited in an extensional basin \citep{Medig2016a}, however it is poorly constrained in terms of age --- older than the Fifteenmile Group and younger than the ca. 1382 Ma Hart River sills (which themselves have been interpreted to be emplaced in conjunction with rifting; \citealp{Verbaas2018a}). As in southwest Laurentia, extensional basin development continued into the Cryogenian and Ediacaran which accommodated deposition of the Windermere Supergroup \citep{Moynihan2019a}. Ediacaran rifting transitioned to thermal subsidence in the Cambrian and an early Paleozoic passive margin \citep{Moynihan2019a}. A similar history is interpreted from the North Slope terrane of Arctic Alaska (considered to have been part of northern Laurentia in the Neoproterozoic) where \cite{Strauss2019a} suggest that there was Tonian-Cryogenian extensional tectonism with limited overall lithospheric stretching that was followed by Ediacaran to early Cambrian extension that led to passive margin sedimentation.

Another margin that experienced rifting and associated passive margin thermal subsidence earlier in the Neoproterozoic is the northeast Greenland margin (Fig. \ref{fig:tectonic_history}). Available geochronological constraints and thermal subsidence modeling indicate ca. 850 to 820 Ma rifting followed by thermal subsidence of a stable platform \citep{Maloof2006a, Halverson2018a}. These data suggest that conjugate continental lithosphere rifted away from northeast Greenland by ca. 820 Ma.

>>>>

The Grenvillian orogeny was a protracted interval of continent-continent collision (ca. 1.09 to 0.98 Ga) leading to amphibolite to granulite facies metamorphism throughout the orogen \citep{McLelland2010a}. Note that while the term Grenville orogeny or Grenville belt is used rather loosely throughout much of the literature to refer to any late Mesoproterozoic orogenic belt, the timeline of orogenesis on the Laurentia margin has more nuanced constraints than this usage. Properly referring to the Grenvillian orogeny as distinct from the Elzevirian and Shawinigan orogenies enables the available constraints to be comparatively assessed when evaluating potential conjugate continents to Laurentia associated with the orogen (Fig. \ref{fig:tectonic_history}). Evidence of large-scale continent-continent collision at the time of the Ottawan Phase of the Grenvillian orogeny is recorded in Texas \citep{Grimes2004a}, through the Blue Ridge Appalachian inliers \citep{Johnson2020a}, through Ontario and to the Labrador Sea \citep{Rivers2008a}. The orogeny is interpreted to have resulted in the development of a thick plateau associated with the Ottawan orogenic phase (ca. 1090 to 1030 Ma; \citealp{Rivers2008a}). Continued convergence during the Rigolet phase of the Grenvillian orogeny led to the development of the Grenville Front tectonic zone and ended ca. 980 Ma \citep{Hynes2010a}.

@@

There is significantly less preserved Mesoproterozoic crustal growth on the western margin of Laurentia (Fig. \ref{fig:Laurentia_map}) and the tectonic history through the Mesoproterozoic Era is not as well elucidated as on the southern to eastern margin. There are thick successions of Paleoproterozoic siliciclastic and carbonate sedimentary rocks such as the ca. 1.66 to 1.62 Ga Wernecke Supergroup in Yukon, Canada \citep{Delaney1986a, Furlanetto2016a}. The Wernecke Supergroup is interpreted to have resulted from rifting followed by passive margin thermal subsidence \citep{Furlanetto2016a}, although the tectonic setting is poorly understood and could be intracratonic. These metasedimentary rocks were deformed and metamorphosed during the ca. 1.60 Ga Racklan-Forward orogeny that is interpreted to be associated with collision of an arc terrane and potentially a conjugate continent \citep{Thorkelson2005a, Furlanetto2013a, Furlanetto2016a}. Further south along Laurentia's western margin, sedimentary rocks of the 15 to 20 km thick Belt-Purcell Supergroup are associated with a ca. 1.47 to 1.40 rift \citep{Evans2000c}. While the rift is typically interpreted as being intracontinental \citep{Lydon2004a}, the tectonic setting in which it formed is debated. \citet{Hoffman1989c} proposed that it may be a remnant back-arc basin trapped within a continent, while others envision it as being associated with continental rifting along the margin associated with separation of a conjugate continent \citep{Jones2015a}. This region is interpreted to have been subsequently deformed during a ca. 1.37 to 1.33 Ga event known as the East Kootenay orogeny that is constrained by granite crystallization and authigenic monazite dates \citep{McMechan1982a, Nesheim2012a, McFarlane2015a}.

Taken together, this late Paleoproterozoic and Mesoproterozoic tectonic history provides significant constraints on paleogeographic reconstructions. In particular, the long-lived history of accretionary orogenesis along the southeast (present-day coordinates) of Laurentia from the initiation of the Yavapai orogeny (ca. 1.71 Ga) to the end of the Shawinigan orogeny (ca. 1.16 Ga) requires a long-lived open margin without a major conjugate continent until the time of terminal Grenvillian collisional orogenesis \citep{Karlstrom2001a}. This constraint is incorporated into paleogeographic models such as that of \citet{Zhang2012a} and \citet{Pehrsson2015a} which maintain a long-lived convergent margin throughout the Mesoproterozoic, but in some reconstructions other continental blocks are reconstructed into positions that are seemingly incompatible with this record of accretionary orogenesis (e.g. Amazonia in \citealp{Elming2009a, Elming2021a}). The high-grade metamorphism associated with the Ottawan phase of the Grenvillian orogeny itself strongly suggests a collision between Laurentia and another continent ca. 1080 Ma --- the geological observation of which first led to the formulation of the hypothesis of the supercontinent Rodinia \citep{Hoffman1991a}. This extensive and major collisional orogenesis on Laurentia's margin, and that of other Proterozoic continents, remains a strong piece of evidence that a supercontinent or (proto)supercontinent formed at the 1.0 Ga Mesoproterozoic to Neoproterozoic transition.

@@

The subsequent Neoproterozoic tectonic history of Laurentia is dominantly a record of rifting (Fig. \ref{fig:tectonic_history}). Along the western margin of Laurentia, by ca. 760 Ma there was rifting leading to deposition in basins from the Death Valley region of southwest Laurentia to the Ogilvie Mountains of northwest Laurentia \citep{Macdonald2013a, Strauss2015a, Dehler2017a, Rooney2017a}. The emplacement of the ca. 780 Ma Gunbarrel large igneous province \citep{Harlan2003a} along this margin and the subsequent extension recorded in the western Laurentia basins is commonly interpreted to be associated with the break-up of Laurentia and a conjugate continent to the western margin (e.g. \citealp{Li2008a}) although this interpretation is difficult to reconcile with the subsequent history of basin development. Extensional basin development continued into the Cryogenian Period with active normal faulting occurring during the deposition of both Sturtian (ca. 717 to 656 Ma) and Marinoan (ca. 645 to 635 Ma) glacial deposits in the Death Valley region of southwest Laurentia \citep{Nelson2020a}. Additionally, Cryogenian volcanics along the western Laurentia margin (e.g. \citealp{Eyster2018a}) are interpreted to be the result of active rifting.  A puzzlingly feature of this record of active rifting is that it significantly predates interpreted passive margin thermal subsidence closer to the ca. 539 Ma Neoproterozoic-Phanerozoic boundary that has been linked to lithospheric thinning \citep{Bond1984a, Levy1991a}. If the interpretation of a conjugate continent rifting of the margin prior to the ca. 717 Tonian-Cryogenian boundary is correct, it is unclear why there would be minimal thermal subsidence until the Ediacaran (after 635 Ma in \citealp{Levy1991a} and \citealp{Witkosky2018a}). While the geological evidence supports prolonged extensional tectonism along the western margin of Laurentia (a portion of which could be strike-slip and transtensional; \citealp{Smith2015b}), it suggests that there was significant lithospheric thinning associated with rifting later than the timing of rifting typically implemented in models of Rodinia break-up. This record of Neoproterozoic basin development led \cite{Yonkee2014a} to propose that the early ca. 780 Ma rifting was intracratonic and that while it may have led to some associated thermal subsidence that there was a second interval of rifting and thermal subsidence associated with Australia rifting away in the Ediacaran (later than in most models). Another possibility, along the lines of that proposed in \citet{Ross1991a} and \citet{Colpron2002a}, is that ca. 760 Ma extensional tectonism is an inboard record of rifting and passive margin development that occurred further to the west. In this model, subsequent continental rifting that drove lithospheric thinning, perhaps associated with the departure of a ribbon continent rather than an already departed major conjugate continent, would be the cause of Ediacaran to Cambrian thermal subsidence.

In northwest Laurentia from the Ogilvie Mountains of Yukon to Victoria Island, the sedimentary rock record is distinct from further south as it also records earlier Neoproterozoic basin development during the Tonian Period in addition to Cryogenian basin development \citep{Macdonald2012a}. Lithospheric extension is interpreted from basin development that accommodated deposition of the lower Fifteenmile Group with maximum depositional ages of ca. 1050 Ma with ongoing basin development ca. 812 Ma (age constraint from a U-Pb zircon date on a tuff within the upper Fifteenmile Group; \citealp{Macdonald2010a}) potentially through thermal subsidence \citep{Macdonald2012a}. Earlier basin development in the region recorded by the Mesoproterozoic/Neoproterozoic Pinguicula Group could provide valuable insight on tectonic history as it has been interpreted to have been deposited in an extensional basin \citep{Medig2016a}, however it is poorly constrained in terms of age --- older than the Fifteenmile Group and younger than the ca. 1382 Ma Hart River sills (which themselves have been interpreted to be emplaced in conjunction with rifting; \citealp{Verbaas2018a}). As in southwest Laurentia, extensional basin development continued into the Cryogenian and Ediacaran which accommodated deposition of the Windermere Supergroup \citep{Moynihan2019a}. Ediacaran rifting transitioned to thermal subsidence in the Cambrian and an early Paleozoic passive margin \citep{Moynihan2019a}. A similar history is interpreted from the North Slope terrane of Arctic Alaska (considered to have been part of northeastern Laurentia in the Neoproterozoic) where \cite{Strauss2019a} interpret that there was Tonian-Cryogenian extensional tectonism with limited overall lithospheric stretching that was followed by Ediacaran to early Cambrian extension that led to passive margin sedimentation.

Another margin that experienced rifting and associated passive margin thermal subsidence earlier in the Neoproterozoic is the northeast Greenland margin, including the displaced terrane of northeastern Svalbard, Norway (Fig. \ref{fig:tectonic_history}). Available geochronological constraints and thermal subsidence modeling indicate ca. 850 to 820 Ma rifting followed by thermal subsidence of a stable platform \citep{Maloof2006a, Halverson2018a}. These data suggest that conjugate continental lithosphere rifted away from northeast Greenland by ca. 820 Ma.
